---
title: "dicepro — Simulated Data Workflow"
author: "dicepro Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{dicepro — Simulated Data Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning    = FALSE,
  message    = FALSE,
  eval       = FALSE
)
```

> **Note:** All code chunks have `eval = FALSE` and are shown for
> illustration only. To run them interactively:
> ```r
> library(dicepro)
> # copy-paste the chunks below into your R session
> ```

# Overview

This vignette demonstrates a complete dicepro deconvolution workflow on
**fully synthetic data**, enabling end-to-end reproducibility without
requiring access to proprietary datasets.

The pipeline proceeds in four stages:

1. **Simulate** a reference signature matrix and ground-truth cell-type
   proportions using the built-in simulation framework.
2. **Generate** a noisy bulk RNA-seq matrix from the simulated mixture.
3. **Deconvolve** using the main `dicepro()` function with automated
   hyperparameter optimisation.
4. **Evaluate** the recovered proportions against the ground truth.

---

# Data Simulation

`simulation()` generates a self-consistent triplet -- reference matrix
`$W`, proportion matrix `$p`, and noisy bulk matrix `$B` -- in a single
call, guaranteeing column-name alignment across all three outputs.

```{r simulate}
library(dicepro)
set.seed(2101L)

sim_data <- simulation(
  loi        = "gauss",
  scenario   = "hierarchical",
  nSample    = 30L,
  nGenes     = 200L,
  nCellsType = 10L,
  sigma_bio  = 0.07,
  sigma_tech = 0.07,
  seed       = 2101L
)

cat("Reference  :", dim(sim_data$W), "\n")
cat("Proportions:", dim(sim_data$p), "\n")
cat("Bulk       :", dim(sim_data$B), "\n")
cat("Row sums   :", round(range(rowSums(sim_data$p)), 4), "\n")
```

```{r sanity-check}
bulk_clean <- as.matrix(sim_data$W) %*% t(as.matrix(sim_data$p))

plot(
  bulk_clean[seq_len(500L)],
  as.matrix(sim_data$B)[seq_len(500L)],
  xlab = "Clean bulk (first 500 entries)",
  ylab = "Noisy bulk",
  pch  = 19, cex = 0.4, col = "#2c7bb6",
  main = "Noise model: clean vs noisy bulk"
)
abline(0, 1, col = "firebrick", lwd = 1.5)
```


# Deconvolution with `dicepro()`

```{r run-dicepro}
out <- dicepro(
  reference             = as.matrix(sim_data$W),
  bulk                  = as.matrix(sim_data$B),
  methodDeconv          = "FARDEEP",
  W_prime               = 0,
  bulkName              = "SimBulk",
  refName               = "SimRef",
  hp_max_evals          = 150L,
  algo_select           = "random",
  output_path           = tempdir(),
  hspaceTechniqueChoose = "all",
  normalize = FALSE
)
```

---

# Results

## Optimal Hyperparameters

```{r best-hp}
cat("Best lambda :", out$hyperparameters$lambda, "\n")
cat("Best gamma  :", out$hyperparameters$gamma,  "\n")
cat("Loss        :", out$metrics$loss,            "\n")
cat("Constraint  :", out$metrics$constraint,      "\n")
```

## Hyperparameter Optimisation Report

```{r plot-hyperopt, fig.height = 9}
out$plot_hyperopt
```

## Pareto Frontier

```{r plot-pareto}
out$plot
```

## Recovered vs True Proportions

```{r compare-props}
true_prop <- as.matrix(sim_data$p)
pred_prop <- as.matrix(out$H)

common_ct   <- intersect(colnames(pred_prop), colnames(true_prop))
true_common <- true_prop[, common_ct, drop = FALSE]
pred_common <- pred_prop[, common_ct, drop = FALSE]

r_overall <- cor(as.vector(true_common), as.vector(pred_common))
cat(sprintf("Overall Pearson r: %.3f\n", r_overall))

plot(
  as.vector(true_common),
  as.vector(pred_common),
  xlab = "True proportions",
  ylab = "Predicted proportions",
  pch  = 19, cex = 0.5, col = "#2c7bb6aa",
  main = sprintf("True vs Predicted  (r = %.3f)", r_overall)
)
abline(0, 1, col = "firebrick", lwd = 1.5)
```

## Per-Cell-Type Correlation

```{r per-ct-cor, fig.height = 4}
ct_cors <- vapply(common_ct, function(ct) {
  cor(true_common[, ct], pred_common[, ct])
}, numeric(1L))

par(mar = c(5, 10, 3, 1))
barplot(
  sort(ct_cors),
  horiz = TRUE, las = 1,
  col   = ifelse(sort(ct_cors) > 0.7, "#2c7bb6", "#d7191c"),
  xlab  = "Pearson r",
  main  = "Per-cell-type correlation",
  xlim  = c(-0.2, 1)
)
abline(v = 0.7, lty = 2, col = "gray40")
```

## Quantitative Performance Metrics

```{r perf-metrics}
perf <- MakeTable1Tool(pred_mat = pred_common, obs_mat = true_common)
knitr::kable(perf$Perf, digits = 3,
             caption = "Performance metrics -- simulated data")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
